\section*{Representing Time}

\subsection*{The representation of time in OWL axioms}

Both the OBO axiom
%
\begin{equation}
\begin{split}
\mclass{CellMembrane}\;\mirel{partOf}\;\mclass{Cell}
\end{split}
\label{eq:cellobo}
\end{equation}  
%
and the equivalent OWL axiom
%
\begin{equation}
\begin{split}
\mclass{CellMembrane}\;\mathtt{subClassOf}\;\mirel{partOf}\;\mathtt{some}\;\mclass{Cell}
\end{split}
\label{eq:celldl}
\end{equation}  
%
are undefined with regard to time. This is not a side aspect, because it makes a
difference whether, e.g., a cell membrane is \emph{always} part of some cell or
only \emph{at some time}.
The lack of temporal definition of OWL statements is especially unsatisfactory
when it comes to transitive properties like $\mirel{partOf}$ or $\mirel{locatedIn}$,
where the suppression of the temporal factor can produce plainly wrong entailments,
especially at the level of individuals, such as in the following example:
%
\begin{equation}
\begin{split}
\mirel{locatedIn} (\mirel{Thrombus\#39874}, \mirel{Heart\#431234})  \\
\mirel{locatedIn} (\mirel{Heart\#431234}, \mirel{Patient\#900812})
\end{split}
\label{eq:trans}
\end{equation}
%
Assuming that the thrombus was no longer in the heart when it was transplanted to Patient\#115678, then the entailment
%
\begin{equation}
\begin{split}
\mirel{locatedIn} (\mirel{Thrombus\#39874}, \mirel{Patient\#115678})
\end{split}
\label{eq:transEntailment}
\end{equation}
%
which follows from the transitivity property of the relation $\mirel{locatedIn}$, is obviously wrong.

This has a far-reaching consequence: by default, OWL statements of this form are temporally ambiguous in a way that may entail unintended reasoning consequences when domain ontologies are used with real data. (A lack of an explicit treatment of temporal context has previously been implicated in an assessment of the quality of existential restrictions in OBO Foundry ontologies \cite{boeker2011}.)

\subsection*{Strengths of Relatedness}

We can make the following distinction between different temporal \emph{strengths} of relatedness.

\subsubsection*{Some Time Relatedness (STR)}

Informally: for all a instances of \mclass{A} there is some time $t$ and some instance $b$ of
\mclass{B} such that $a$ is related to $b$ at $t$. Examples:
\begin{enumerate}[(a)]
\item for all apple seeds there is
some apple such that the seed is part of the apple at some time.
\item for all
trees there is some leaf such that the leaf is part of the tree at some time.
\end{enumerate}

Formally:
\begin{equation}
\begin{split}
\mclass{TemporarilyRelated}(\mclass{A},\mclass{B}) =_{def}&\;
\forall a, t:\; \mrelt{inst}(\mclass{A}, a, t) \\
&\ \rightarrow
\exists b, t^\prime:\;(\mrelt{inst}(\mclass{B},b,t^\prime) \wedge
\mrelt{rel}(a,b,t^\prime) \wedge \mirel{within}(t^\prime,t))
\end{split}
\label{eq:temporarily:cls}
\end{equation}

\subsubsection*{Permanent Generic Relatedness (PGR)}

Informally: for all instances $a$ of \mclass{A} there is, at all times $t$ that
$a$ exists,
some instance $b$ of \mclass{B} such that $a$ is related to $b$ at $t$, but not necessarily
always the same $b$ at all times $t$. Examples:
\begin{enumerate}[(a)]
\item all cells have a water molecule as
part at all times, but not always the same water molecule.
\item every bacteria colony has some bacteria as parts at all times, but not
always the same bacteria.
\end{enumerate}

\begin{equation}
\begin{split}
\mclass{PermanentlyGenericallyRelated}(\mclass{A},\mclass{B}) =_{def}&\;
\forall a, t:\; \mrelt{inst}(\mclass{A}, a, t) \\
&\ \rightarrow
\exists b:\;(\mrelt{inst}(\mclass{B},b,t) \wedge
\mrelt{rel}(a,b,t))
\end{split}
\label{eq:generically:cls}
\end{equation}


\subsubsection*{Permanent Specific Relatedness (PSR)}

Informally, for all instances $a$ of \mclass{A} there is, at all times $t$ that $a$ exists, an
instance $b$ of \mclass{B} such that $a$ is related to $b$ at $t$; in this case it is always the
same $b$ at all times $t$. Examples:
\begin{enumerate}[(a)]
\item a human being has a brain as part at all times, and it is necessarily the same brain.
\item a radioactively marked molecule of DNA has the radioactive isotope as part
at all times, and it is necessarily the same atom.
\end{enumerate}

\begin{equation}
\begin{split}
\mclass{Permanently}&\mclass{SpecificallyRelated}(\mclass{A},\mclass{B}) =_{def}\;
\forall a, t:\; \mrelt{inst}(\mclass{A}, a, t) \\
&\ \rightarrow
\exists b:\;\big(\mrelt{inst}(\mclass{B},b,t) \wedge
\mrelt{rel}(a,b,t))
\\
&\quad\quad \wedge \forall t^\prime: (\mrelt{inst}(\mclass{A},a,t^\prime)
\rightarrow (\mrelt{rel}(a,b,t^\prime) \wedge
\mrelt{inst}(\mclass{B},b,t^\prime))\big)
\end{split}
\label{eq:specifically:cls}
\end{equation}


\subsection*{Use Cases and Competency Questions}

\todo[inline, size=\small]{Shouldn't the use cases go into the Methods section?}


\todo[inline, size=\small]{Possible competency questions: (i) phased sortals, (ii) changing location at individual level (iii) class-level queries on anatomical objects (iv) class-level queries in which other than mereological relations occur (e.g. inherence, participation)}

The following use cases (UC) and competency questions (CQ) will be used for a comparison of the two representational approaches we are proposing. 

The use cases will be structured as follows: For each temporal mode of existence (STR, PGR, PSR) we will have four use cases: 
class-level (c), individual-level (i) mereologic (m+), non-mereologic (m-), participation (p), with quantification over the participant (p1) and over the process (p2). In addition we will have a class-level and an individual-level use case for a non-rigid class. 

\begin{itemize}
\item STR\_cm+: Every tooth is part of an organism at some time. Challenged by: \\ ``tooth that is never part of an organism'': insatisfiable \\
``tooth that is part of something that is not an organism or organism part'': satisfiable 


\item STR\_cm-: Every apple is green at some time
\\
Challenged by: \\ ``apple that is never green'': insatisfiable \\
``apple that is red at all times'': insatisfiable 


\item STR\_cp1: Every vertebrate participates in some birth process at some time\\
Challenged by: \\ ``Vertebrate that never participates in any birth process'': insatisfiable \\
``Vertebrate that at some time does not participate in a birth process'': insatisfiable 



\item STR\_cp2: Every fecundation has some spermatozoon as participant at some time
\\
Challenged by: \\ ``Fecundation event for which there is never any spermatozoon that participates'': insatisfiable \\
``Phase of fecundation in which there is no spermatozoon participating'': satisfiable 




\item STR\_im+: The blood specimen \#bs430912 is in the lab \#lab996 at time \#20131020

Challenged by: \\ ``The blood specimen \#bs430912 is in the lab \#lab100 at time \#20131020. Both labs do not share parts'' ==> unsatifiable



\item STR\_im-: Joe's left ankle is swollen at time \#20131020

Challenged by: \\ ``Joe's left ankle is not swollen at time \#20131021'' ==> satifiable
\\ ``Joe's left ankle at time \#20131020 is not swollen'' ==> unsatifiable



\item STR\_ip:  Mary participated in the process Mary's birth

Challenged by: \\``There is a time in which Mary is not a participant of her birth''  ==> satifiable

%%%%%%%


\item PGR\_cm+: Every red blood cell always includes some oxygen molecule 







Challenged by: \\``There are red blood cells without oxygen molecules''  ==> unsatifiable


\item PGR\_cm-: Every electronic health record is always on some computer storage

Challenged by: \\``There are electronic health record that are not on a computer storage''  ==> unsatifiable


\item PGR\_cp1: Every ?  \todo{any idea?}
\item PGR\_cp2: Every breathing process has always some volume of air as participant 

Challenged by: \\``There are breathing processes in which no air volume participates''  ==> unsatifiable


\item PGR\_im+: Mary's blood always consists of some blood cell 

Challenged by: \\``There is a time in which Mary's blood is devoid of blood cells''  ==> unsatifiable

\item PGR\_im-: Joe's electronic health record always resides on some electronic storage medium

Challenged by: \\``There is a time in which Joe's electronic health is not on an electronic storage''  ==> unsatifiable


\item PGR\_ip:  Joe's breathing process has always some volume of air as participant

Challenged by: \\``Joe's breathing process is without air at some time''  ==> unsatifiable

%%%%%%%%%%%

\item PSR\_cm+: Every human has a brain and it is always the same brain

Challenged by: \\``A human who has more than one brain''  ==> unsatifiable

\todo[inline, size=\small]{``This is rather tricky with TQC. One could express it as: for all temporal 
instances of Human there is always some temporal instance of brain related. For all human instance which has the same "max" 
all related brain instances must also have the same "max" ''}





\item PSR\_cm-: Every mammal has the same sex during its life

Challenged by: \\``A mammal that has more than one sex''  ==> unsatifiable

 
\item PSR\_cp1: Every organism always participates in its life, and it is always the same

Challenged by: \\``An  organism that participates in two different lifes''  ==> unsatifiable


\item PSR\_cp2: Every biological life always has some organism as participant, and it is always the same

Challenged by: \\``A life that has more than one organism as participant''  ==> unsatifiable


\item PSR\_im+: Mary has `Mary's brain' as a proper part during her lifetime 

Challenged by: \\``Mary gets a new brain during her lifetime''  ==> unsatifiable


\item PSR\_im-: Tibbles has the same sex during his life 

Challenged by: \\``Tibbles changes his sex during his life''  ==> unsatifiable


\item PSR\_ip: Joe participates in his life and not in any other life

Challenged by: \\``Joe participates in more than one life''  ==> unsatifiable


%%%%%%%%%%%


\item NR\_t: Every fetus is an embryo at some time. Nothing can be both an embryo and a fetus at the same time. Vertebrate embryoes have a chorda at some time

Challenged by: \\ ``There are fetuses that were never embryoes''  ==> unsatifiable




\item NR\_i: John was a medical student from 2001-10-01 to 2007-06-30

Challenged by: \\ ``Every person which is a medical student was a high school student at some time''   
John was never a high school student. ==> unsatisfiable




For the implementation of the use cases and competency questions as OWL axioms see supplementary data.
 

Development process of gastrulation which starts with one anatomical part and goes on to another anatomical part? See gastrulation wikipedia page. 
I see this part. Which phases/stages could my organism be in?

During neural development the cells are doing something specific. Existence of process defines phase.

\todo[inline, size=\small]{Alan has a worked out example of the phases of life of some moth organism.}
\todo[inline, size=\small]{Embryo has this part but the fetus does not have that part.}
\todo[inline, size=\small]{Joe's income when he was a medical student was lower than his income after he graduated.}
\todo[inline, size=\small]{An assertion about joe as a medical student shouldn't make it on to Joe in general.}

We will use these use cases and competency questions to evaluate two proposed design patterns for temporalization against the following functional criteria:
\begin{enumerate}
    \item technical completeness, 
    \item user friendliness, and
    \item performance profile
\end{enumerate}
The details of how we conducted the evaluation are described in our Methods section.

\end{itemize}




 
